---
title: "Introduction to Bayesian Methods"
subtitle: 'Activity 2A: Analytic vs. Numeric Bayesian analysis'
format:
  html:
    toc: true
    toc-location: left
    html-math-method: katex
    css: styles.css
---

[Main materials](materials.qmd)

# Introduction

This section is focused on practicing the basics of Bayesian analysis (both analytic practice and computation using nimble) for simple uni-modal data. This section assumes that you have seen both of the Bayes lectures ([here](VB_Bayes1.qmd), and [here](VB_Bayes2.qmd)). We'll review some of the analytic results so that you can compare the results from the previous activity to numerical results while also learning about how to check your computation. We suggest some more advanced things to try for those interested in implementing your own models.

## Packages and tools

For this practical you will need to first install [nimble](), then be sure to install the following packages:

```{r, results="hide", warning=FALSE, message=FALSE}
# Load libraries
require(nimble)
require(HDInterval)
library(MCMCvis)
require(coda) # makes diagnostic plots
require(IDPmisc) # makes nice colored pairs plots to look at joint posteriors

##require(mcmcplots) # another option for diagnostic plots, currently unused
```

# Fitting Models the Bayesian way

Recall from the lectures (see above for links!) that in order to make inference statements (for example, credible intervals) that we need to either 1) known the full form of the posterior distribution or 2) have numerical samples from the target distribution. For the second option, we need the samples to be **good** -- that is they need to be samples from the posterior and we need the samples in the Markov Chain to be not too highly correlated with each other. Thus our numerical steps in our analysis would be as follows:

1.  **Implement and fit the model** using your tools of choice

2.  **Assess MCMC convergence**: MCMC is family of algorithms for sampling from probability distributions (e.g., a posterior distribution in a Bayesian analysis). The MCMC procedure reaches *convergence* once the algorithm is taking draws from the posterior and not still trying to figure it out. To assess convergence our primary tool is to we look at trace plots (i.e., parameter samples taken in order, like a time series). The goal is to get traces that look like "fuzzy caterpillars".

3.  **Summarize MCMC draws**: Summarize and visualize outcome of the random draws using histograms for all draws for each parameter, and calculate expectation, variance, credibility interval, etc.

4.  **Prior Sensitivity**: Assess prior sensitivity by changing prior values and checking the extent to which it affects the results. If it does, that means that the results may be too sensitive to that prior, usually not good!

5.  **Make inferences**: We use the values from item (2) to make inferences and answer the research question.

We'll review the example data, and the analytic likelihoods before moving on to the computational approximation of the posterior using nimble, and go through each of these steps.

## Example: Midge Wing Length

We will use this simple example to go through the steps of assessing a Bayesian model and we'll see that MCMC can allow us to approximate the posterior distribution.

Grogan and Wirth (1981) provide data on the wing length (in millimeters) of nine members of a species of midge (small, two-winged flies).

From these measurements we wish to make inference about the population mean $\mu$.

```{r, fig.align='center'}
# Load data
WL.data <- read.csv("MidgeWingLength.csv")
Y <- WL.data$WingLength
n <- length(Y)

hist(Y,breaks=10,xlab="Wing Length (mm)") 
```

We'll also need summary statistics for the data that we calculated last time:

```{r}
m<-sum(Y)/n
s2<-sum((Y-m)^2)/(n-1)
```

## Recall: Setting up the Bayesian Model

We need to define the likelihood and the priors for our Bayesian analysis. Given the analysis that we've just done, let's assume that our data come from a normal distribution with unknown mean, $\mu$ but that we know the variance is $\sigma^2 = 0.025$. That is: $$
\mathbf{Y} \stackrel{\mathrm{iid}}{\sim} \mathcal{N}(\mu, 0.025^2)
$$

In the last activity we our prior for $\mu$ to be be: $$
\mu \sim \mathcal{N}(1.9, 0.8^2)
$$ Together, then, our full model is: $$
\begin{align*}
\mathbf{Y} & \stackrel{\mathrm{iid}}{\sim} \mathcal{N}(\mu, 0.025^2)\\
\mu &\sim \mathcal{N}(1.9, 0.8^2)
\end{align*}
$$

In the previous activity we wrote a function to calculate $\mu_p$ and $\tau_p$ and then plugged in our numbers:

```{r}
tau.post<-function(tau, tau0, n){n*tau + tau0}
mu.post<-function(Ybar, mu0, sig20, sig2, n){
  weight<-sig2+n*sig20
  
  return(n*sig20*Ybar/weight + sig2*mu0/weight)
}
```

Finally we plotted 3 things together -- the data histogram, the prior, and the posterior

```{r}
mu0 <- 1.9
s20 <- 0.8
s2<- 0.025 ## "true" variance

mp<-mu.post(Ybar=m, mu0=mu0, sig20=s20, sig2=s2, n=n)
tp<-tau.post(tau=1/s2, tau0=1/s20, n=n)
```

```{r, fig.align='center'}
x<-seq(1.3,2.3, length=1000)
hist(Y,breaks=10,xlab="Wing Length (mm)", xlim=c(1.3, 2.3),
     freq=FALSE, ylim=c(0,8)) 
lines(x, dnorm(x, mean=mu0, sd=sqrt(s20)), col=2, lty=2, lwd=2) ## prior
lines(x, dnorm(x, mean=mp, sd=sqrt(1/tp)), col=4, lwd=2) ## posterior
legend("topleft", legend=c("prior", "posterior"), col=c(2,4), lty=c(2,1), lwd=2)
```

# Numerical evaluation of the posterior with nimble

Let's show that we can get the same thing from nimble that we were able to get from the analytic results. You'll need to make sure you have installed nimble (which must be done outside of R) and then the R libraries ${\tt coda}$ and ${\tt nimble}$.

## Specifying the model

First we must encode our choices for our data model and priors to pass them to the fitting routines in nimble. This involves setting up a ${\tt model}$ that includes the likelihood for each data point and a prior for every parameter we want to estimate. Here is an example of how we would do this for the simple model we fit for the midge data (note that nimble uses the precision instead of the variance or sd for the normal distribution):

```{r}
modelCode <-  nimbleCode({

  ## Likelihood
  for(i in 1:n){
    Y[i] ~ dnorm(mu,tau)
  }

  ## Prior for mu
  mu  ~ dnorm(mu0,tau0)

} ## close model
)
```

This model is formally in the BUGS language (also used by JAGS, WinBugs, etc). Now we will create the nimble model

```{r}

model1 <- nimbleModel(code = modelCode, name = "model1", 
                    constants = list(tau=1/s2, mu0=mu0,
                                    tau0=1/s20, n=n),
                    data  = list(Y=Y), 
                    inits = list(mu=5))

model1$getNodeNames()

## just checking it can compile
Cmodel1<- compileNimble(model1)

```

Then we run the MCMC and, see how the output looks for a short chain:

```{r}

mcmc.out <- nimbleMCMC(code = modelCode, 
                       constants = list(tau=1/s2, mu0=mu0,
                                    tau0=1/s20,n=n),
                       data  = list(Y=Y),
                       inits = list(mu=5),
                       nchains = 1, niter = 100,
                       #summary = TRUE, WAIC = TRUE,
                       monitors = c('mu'))

dim(mcmc.out)
head(mcmc.out)

```

```{r, fig.align='center'}
samps<-as.mcmc(mcmc.out)
plot(samps)
```

MCMC is a rejection algorithm that often needs to converge or "burn-in" -- that is we need to potentially move until we're taking draws from the correct distribution. Unlike for optimization problems, this does not mean that the algorithm heads toward a single value. Instead we're looking for a pattern where the draws are seemingly unrelated and random. To assess convergence we look at trace plots, the goal is to get traces that look like "fuzzy caterpillars".

Sometimes at the beginning of a run, if we start far from the area near the posterior mean of the parameter, we will instead get something that looks like a trending time series. If this is the case we have to drop the samples that were taken during the burn-in phase. Here's an example of how to do that, also now running 2 chains simultaneously.

```{r}
mcmc.out <- nimbleMCMC(code = modelCode, 
                       constants = list(tau=1/s2, mu0=mu0,
                                    tau0=1/s20,n=n),
                       data  = list(Y=Y),
                       inits = list(mu=5),
                       nchains = 2, niter = 11000,
                       nburnin = 1000,
                       #summary = TRUE, WAIC = TRUE,
                       monitors = c('mu'))

dim(mcmc.out$chain1)
head(mcmc.out$chain1)
```

```{r, fig.align='center'}
samp<-as.mcmc(mcmc.out$chain1)
plot(samp)
```

This is a very fuzzy caterpillar!

We also often want to check the autocorrelation in the chain.

```{r}
acfplot(samp, lag=20, aspect="fill", ylim=c(-1,1))
```

This is really good! It means that the samples are almost entirely uncorrelated.

Finally we can also use the summary function to examine the samples generated:

```{r}
summary(samp)
```

Let's compare these draws to what we got with our analytic solution:

```{r, fig.align='center'}
x<-seq(1.3,2.3, length=1000)
hist(samp, xlab="mu", xlim=c(1.3, 2.3),
     freq=FALSE, ylim=c(0,8), main ="posterior samples") 
lines(x, dnorm(x, mean=mu0, sd=sqrt(s20)), col=2, lty=2, lwd=2) ## prior
lines(x, dnorm(x, mean=mp, sd=sqrt(1/tp)), col=4, lwd=2) ## posterior
legend("topleft", legend=c("prior", "analytic posterior"), col=c(2,4), lty=c(2,1), lwd=2)
```

It worked!

As with the analytic approach, it's always a good idea when you run your analyses to see how sensitive is your result to the priors you choose. Unless you are purposefully choosing an informative prior, we usually want the prior and posterior to look different, as we see here. You can experiment yourself and try changing the prior to see how this effects the posterior.

<br> <br> <br>

## Practice: Applying to a new dataset

Download VecTraits dataset 562 (Kutcherov et al. 2018. Effects of temperature and photoperiod on the immature development in Cassida rubiginosa Mull. and C. stigmatica Sffr. (Coleoptera: Chrysomelidae). Sci. Rep. 9: 10047). This dataset explores the effects of temperature and photoperiod on body size (weight in mg) for *Cassida stigmatica*, a type of small beetle. Subset the data so you focus on one temperature/photoperiod combination (you could also choose to subset by sex). Using the same model as above, potentially with a different prior distribution, and with the value of $\tau$ set to $1/s^2$ (where $s$ is the empirical standard deviation of your dataset), redo the analysis above.

<br> <br> <br>

# Optional Advanced Exercise: Estimating the population variance

One advantage of the numerical approach is that we can choose almost anything we want for the priors on multiple parameters without worrying if they are conjugate, or if we want to include additional information. For example, let's say that we want to force the mean to be positive (and also the data, perhaps), and concurrently estimate the variance. Here is a possible model:

```{r}
modelCode2 <-  nimbleCode({

  ## Likelihood
  for(i in 1:n){
    Y[i] ~ T(dnorm(mu,tau),0, ) ## truncates at 0
  }

  ## Prior for mu
  mu  ~ T(dnorm(mu0,tau0), 0,)
  
  # Prior for the precision
  tau   ~ dgamma(a, b)

  # Compute the variance
  s2  <- 1/tau

} ## close model
)

```

```{r}
## hyperparams for tau
a   <- 0.01
b   <- 0.01

## alternative approach -- save constants, data, etc, 
## as lists outside the function

modConsts2 <- list(mu0=mu0,tau0=1/s20, a=a, b=b, n=n)
modData2   <- list(Y=Y)
modInits2  <- list(mu=5, tau=1/s2)

model2 <- nimbleModel(code = modelCode2, name = "model2", 
                    constants = modConsts2,
                    data  = modData2, 
                    inits = modInits2)

model2$getNodeNames()

## just checking it can compile
Cmodel2<- compileNimble(model2)


```

```{r}
mcmc.out2 <- nimbleMCMC(code = modelCode2, 
                       constants = modConsts2,
                       data  = modData2, 
                       inits = modInits2,
                       nchains = 2, niter = 11000,
                       nburnin = 1000,
                       #summary = TRUE, WAIC = TRUE,
                       monitors = c('mu', 's2', 'tau'))

dim(mcmc.out2$chain1)
head(mcmc.out2$chain1)

samp2<-as.mcmc(mcmc.out2$chain1)
```

```{r, fig.align='center'}

plot(samp2)
```

```{r}
summary(samp2)
```

Now we plot each with their priors:

```{r, fig.align='center'}
par(mfrow=c(1,2), bty="n")

hist(samp2[,1], xlab="samples of mu", main="mu", freq=FALSE)
lines(x, dnorm(x, mean=mu0, sd=sqrt(s20)), 
      col=2, lty=2, lwd=2) ## prior

x2<-seq(0, 200, length=1000)
hist(samp2[,3], xlab="samples of tau", main="tau", freq=FALSE)
lines(x2, dgamma(x2, shape = a, rate = b), 
      col=2, lty=2, lwd=2) ## prior

```

We also want to look at the joint distribution of $\mu$ and $\sigma^2$:

```{r, fig.align='center'}
plot(as.numeric(samp2[,1]), as.numeric(samp2[,2]), 
     xlab="mu", ylab="s2")
```

Or, a prettier plot:

```{r}
ipairs(as.matrix(samp2[,1:2]), ztransf = function(x){x[x<1] <- 1; log2(x)})
```

## Practice: Updating the model

Redo the previous analysis placing a gamma prior on $\mu$ as well. Set the prior so that the mean and variance are the same as in the normal example from above (use moment matching). Do you get something similar?

<br> <br> <br>
